---
title: "Generate sequential patterns"
author: "Nova F. Smedley"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
vignette: >
  %\VignetteIndexEntry{Generate sequential patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package `gbmSPM` can be used to do the exact analysis pipeline used in the paper, 

Smedley, Nova F., Benjamin M. Ellingson, Timothy F. Cloughesy, and William Hsu. "Longitudinal Patterns in Clinical and Imaging Measurements Predict Residual Survival in Glioblastoma Patients." *Scientific reports* 8, no. 1 (2018): 14429.

See also Supplemental Materials.

## Vignette Info

This vignette shows how temporal patterns can be found with sequential pattern mining (SPM) using dummy patient data and depends on the `arulesSequences` R package.


## Example
1. **Set the `arulesSequences` parameters**

    * `minSupp` min. percentage of patients, i.e., sequences, with pattern
    * `maxgap` max. time between CONSECUTIVE elements in sequence, i.e., max. days between visits 
    * `maxlen` max. length of elements in sequence
    * `maxsize` max. number of items in an element of a sequence



    E.g., if min. support is 0.4, only patterns that occur in 40% of patients will be returned.

    E.g., if max. length is 2, max. gap is 60, patterns are created where there is a sequence of up to 2 visits, with up to 60 days between consecutive visits

    E.g., if max. size is 2, there can be up to two events included in a single visit.

    ```{r, message = FALSE}
    library(gbmSpm)

    minSupp <- 0.4
    maxgap <- 60
    maxlen <- 2
    maxsize <- 2
    ```

1. **Set the gbmSPM parameters**

    * `tType` the type of tumor volume measurment to use in generating patterns
    * `chemoOverlap` whether to only use events during the course of chemotherapy (only really relevant for descriptive statistics)
    * `suppList` can subset other sets patterns based on level of min. support specified in this list (reduces duplicated work)

    E.g., if `tType` is 'rate', rate changes will be calculated from tumor volumes and used

    ```{r}
    tType <- 'rate'
    chemoOverlap <- F
    suppList <- seq(minSupp, 0.4, .05)
    ```


1. **Setup experiment to save generated patterns and metadata**

    We wil put the created features in '~/gbm_spm_example', but this can be changed.

    ```{r, message = FALSE, results = "hide"}
    outputDir <- '~/gbm_spm_example'
    spmDir <- file.path(outputDir,'spm')
    logDir <- file.path(outputDir,'spm_logs')
    lapply(c(outputDir, spmDir, logDir), function(i) ifelse(!dir.exists(i), dir.create(i, showWarnings = F, recursive = T), F) )
    ```

    Log experiment:

    ```{r, message = FALSE,}
    # if you want to save the log file, uncomment
    # logfn <- file(file.path(logDir,paste0(paste0('g',maxgap,'l',maxlen,'z',maxsize),'_',
    #                                       getVolTypeName(tType),'.log')), open='wt')
    #
    # sink(logfn, type='output', split=T)

    cat(format(Sys.Date(), format="%Y.%m.%d"), '\n')
    cat('maxgap: ', maxgap, '\n')
    cat('maxlen: ', maxlen, '\n')
    cat('maxsize: ', maxsize, '\n')
    cat('tumor vol variable type: ', getVolTypeName(tType), '\n\n')
    ```

1. **Load dummy data**

    ```{r}
    data("fake_data")
    names(fake_data)
    colnames(fake_data$person)
    colnames(fake_data$demo)
    colnames(fake_data$events)

    # save tumor location and laterility strings before event cleaning
    fake_tumorInfo <- fake_data$events
    fake_demo <- fake_data$demo

    fake_data$events <- cleanData(fake_data$events, tType)

    cat('...',nrow(fake_data$events), " events left for SPM after cleaning", '\n')
    ```

1. **Prep data by collecting patient info for each event**

    ```{r}
    fake_data <- merge(fake_data$events, fake_data$person, by='iois', all.x=T)
    colnames(fake_data)

    # prep for each event, since age does change
    fake_data <- prepDemographics(fake_data, fake_demo)
    colnames(fake_data)

    # get survival labels, these also change
    fake_data <- prepSurvivalLabels(fake_data)
    colnames(fake_data)

    # get first tumor location
    fake_data <- getTumorLocation(fake_data, fake_tumorInfo)
    colnames(fake_data)

    ```

1. **Run spm**

    If you want to save the patterns and use in logit modeling:

    ```{r}
    runSPM(event = fake_data,
           suppList = suppList,
           maxgap = maxgap,
           maxlen = maxlen,
           maxsize = maxsize,
           tType = tType,
           outputDir = spmDir)
    ```

    otherwise:

    ```{r}
    # spm
    pSPM <- getSeqPatterns(event = fake_data,
                           transFilename = file.path(outputDir, 'example_transactions.txt'),
                           createT = T,
                           support = minSupp,
                           maxgap = maxgap,
                           maxlen = maxlen,
                           maxsize = maxsize)
    pSPM$patterns <- as(pSPM$freqseq, "data.frame")
    pSPM$patterns$sequence <- as.character(pSPM$patterns$sequence)

    # days when pattern occur
    patternDays <- findPatternDays(pSPM$patterns, pSPM$data, maxgap=maxgap)

    # feature vectors to supply to logits
    feat_vecs <- getFeatureVectors(patternDays, events=pSPM$data)
    ```

