---
title: "Model selection and plotting performance"
author: "Nova F. Smedley"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Model selection and plotting performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = 'png'
)
```


The package `gbmSPM` can be used to do the exact analysis pipeline used in the paper, 

Smedley, Nova F., Benjamin M. Ellingson, Timothy F. Cloughesy, and William Hsu. "Longitudinal Patterns in Clinical and Imaging Measurements Predict Residual Survival in Glioblastoma Patients." *Scientific reports* 8, no. 1 (2018): 14429.


See also Supplemental Materials.

## Vignette Info

This vignette shows how temporal features and patient covariates are used in predicting residual survival. This uses dummy patient data and follows the other vignettes "Generate sequential patterns" and "Predicting residual survival."

## Example

1. **Create a few types of temporal features**
    
    To use dummy data, run `example_spm.R ` from the `examples` folder in the terminal:

    ```
    $ Rscript /PATHTO/example_spm.R --tType 'rate' --minSupp 0.2 --minSuppList 'yes' --maxgap 60 --maxlength 2 --maxsize 2 --outDir ~/gbm_spm_example_multi
    ```
    
    Optionally: repeat the above for `tType` of `p` for percent change instead of rate change in tumor volume measurements:
    
    ```
    $ Rscript /PATHTO/example_spm.R --tType 'p' --minSupp 0.2 --minSuppList 'yes' --maxgap 60 --maxlength 2 --maxsize 2 --outDir ~/gbm_spm_example_multi
    ```

    This will generate patterns and feature vectors stored in `gbm_spm_example_multi`.
    
1. **Use the generated features in predicting residual survival**
    
    
    To use dummy data, run `example_logit_spm.R ` from the `examples` folder for the different SPM parameters:

    ```
    $ Rscript example_logit_spm.R --tType 'rate' --maxgap 60 --maxlength 2 --lmax 0.2 --llength 100 --dataFolder sup0.4g60l2z2 --prefix logits --dir ~/gbm_spm_example_multi --saveLogit yes
    ```
    
    Optionally: again, repeat the above for the different SPM patterns created:
    * `sup0.2g60l2z2`
    * `sup0.25g60l2z2`
    * `sup0.3g60l2z2`
    * `sup0.35g60l2z2`
    * `sup0.4g60l2z2`
    
    and/or also repeat for another `tType` of `p` for percent change.
  
    This will perform repeated 2-fold cross-validation for hyperparameter selection. Experiment logs and logit model (and results) are also stored in `gbm_spm_example_multi`.  
    
1. **Load functions to format logit results**

    These are functions for processing the data, which were stored and named in a specific manner that would likely be different than how users may use `gbmSPM`.
    
    ```{r,  message = FALSE}
    library(gbmSpm)
    library(grid)
    library(scales)
    library(wesanderson) 
    
    getCVResults <- function(dir, fn, name) {
      cvResultsFile <- file.path(dir,fn)
      cvResults <- read.table(cvResultsFile, header=T, sep=',')
      
      # identify the cspade parameters the logit's features were based on
      temp <- do.call(rbind, lapply(cvResults$model, function(n) regmatches(n, gregexpr('[0-9]+',n))[[1]][2:5] ) )
      colnames(temp) <- c('support', 'gap', 'length', 'size')
      temp <- as.data.frame(temp)
      temp$support <- paste0( '.', as.character(temp$support))
      cvResults <- cbind(cvResults, temp)
      
      cvResults <- cvResults[order(cvResults$model),]
      cvResults$name <- rep(name, nrow(cvResults))
      
      ind <- length(strsplit(as.character(cvResults$model[1]), '\\/')[[1]])
      temp <- strsplit(as.character(cvResults$model),'\\/')
      cvResults$model <- unlist(lapply(temp, '[[', ind))
      cat('duplicates?: ', which(duplicated((cvResults))) )
      return(cvResults)
    }
    
    getCVByMonth <- function(cvResults, month, metricColName) {
      cv <- cvResults[cvResults$month==month,]
      cv <- cv[order(-cv[,metricColName]),]
      cv$model <- factor(cv$model, levels=unique(cv$model)) # order levels by decreasing ROC
      cv$name <- as.factor(cv$name)
      cv$group <- paste0(cv$tType, cv$name)
      cv$tType <- as.factor(cv$tType)
      cv$tType <- factor(cv$tType, levels=levels(cv$tType)[c(3,1,2,4)])
      cv
    }
    
    rankModels <- function(cvsResults, months, topN, metricColName) {
      r <- lapply(months, function(m) {
        models <- unique(cvsResults$model)
        ranked <- lapply(models, function(m) {
          results <- max(cvsResults[cvsResults$model==m,metricColName])
        })
        names(ranked) <- models
        ranked <- unlist(ranked)
        ranked <- ranked[order(-ranked)]
        names(ranked)[1:topN]
      })
      r <- unlist(r)
      r <- r[!duplicated(r)]
      return(r)
    }
    ```
    
    ROC and PR plotting functions:
    
    ```{r}
    # performance plot constants
    perfTheme <- theme(legend.text=element_text(size=8, color="grey30"),
                       legend.title=element_text(size=8), 
                       legend.text.align =1,
                       legend.position='right', 
                       legend.key.size=unit(.8,'line'),
                       legend.margin=margin(l = -0.15, unit='cm'),
                       legend.background = element_rect(fill = "transparent", colour = "transparent"),
                       axis.text=element_text(size=10, color="#666666"),
                       axis.title.y = element_text(margin=margin(0,5,0,0),size=8, angle=0, vjust=0.5),
                       axis.title.x = element_text(margin=margin(5,0,-5,0),size=8),
                       panel.spacing = unit(.35, "lines"),
                       strip.text.x = element_text(size=9,margin = margin(.05,0,.05,0, "cm")))
    scaleX <- scale_x_continuous(limits=c(0,1),breaks=seq(0,1,0.2))
    scaleY <- scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2))
    perfGuide <- guides(colour = guide_legend(override.aes = list(size=3)))
    
    # see https://learnr.wordpress.com/2009/04/29/ggplot2-labelling-data-series-and-adding-a-data-table/
    
    vplayout <- function(...) {
      grid.newpage()
      pushViewport(viewport(layout = Layout))
    }
    subplot <- function(x, y) viewport(layout.pos.row = x,
                                       layout.pos.col = y)
    mmplot <- function(a, b) {
      vplayout()
      print(a, vp = subplot(1, 1))
      print(b, vp = subplot(2, 1))
    }
    # end
    Layout <- grid.layout(nrow = 2, ncol = 1, heights = unit(c(2,0.5), c("null", "null")))
    pd <- position_dodge(0.6) # move them .05 to the left and right

    # ROC
    plotCVResults <- function(cv, month, filename) {
      g <- ggplot(cv, aes(x=model, y=rocAUC, colour=tType, line=tType, group=group) ) +
        geom_point() + geom_line() +
        scale_y_continuous('Average\nArea Under the ROC Curve\n',limits=c(0.4,0.7),breaks=seq(0.4,0.7,.05), minor_breaks=seq(0.4,0.7,0.01)) +
        theme_linedraw() + 
        ggtitle(month) +
        theme(axis.text.x=element_blank(),
              axis.title.x=element_blank(), 
              axis.text=element_text(size=8),
              axis.title.y = element_text(angle = 90, vjust = .5, size=8, margin=margin(0,10,0,0)),
              legend.text=element_text(size=8, color='grey30'), 
              legend.title=element_text(size=8),
              legend.position = c(0.7,0.8),
              legend.key.size=unit(.8,'line'),
              plot.title = element_text(size=8, hjust=0.5)) +  
        labs(color='tumor volume type') +
        scale_color_manual(values = wes_palette("Darjeeling1")[-4] )
      
      table <- cv[!(duplicated(cv$model)),]
      table <- table[, c('model','size','length','gap','support')]
      table <- melt(table, id='model')
      
      t <- ggplot(table, aes(x=model,y=variable, label=format(value))) +
        geom_text(size = 4.2, color='grey30', hjust=0.6) + theme_bw() + scale_y_discrete('cSPADE\nparameters') +  scale_x_discrete(expand=c(0.1,0)) +
        theme(panel.grid.major = element_blank(), legend.position = "none",
              panel.border = element_blank(),
              axis.text.x = element_blank(),
              axis.ticks = element_blank(),
              axis.text=element_text(size=8),
              axis.title.y = element_text(angle = 90, vjust = .5, size=8, margin=margin(0,11,0,5)) ) +
        theme(plot.margin = unit(c(0,.19, 0, 0), "lines")) + xlab(NULL) + ylab(NULL)
      
      a <- mmplot(g, t)
    }
    
    # PR
    plotCVResults_PR <- function(cv, month, filename, lp) {
      g <- ggplot(cv, aes(x=model, y=prAUC, colour=tType, line=tType, group=group) ) +
        geom_point() + geom_line() +
        scale_y_continuous('Average\nArea Under the PR Curve\n',limits=c(0,0.2),breaks=seq(0,0.2,.05), minor_breaks=seq(0,0.2,0.01)) +
        theme_linedraw() + 
        ggtitle(month) +
        theme(axis.text.x=element_blank(),
              axis.title.x=element_blank(), 
              axis.text=element_text(size=8),
              axis.title.y = element_text(angle = 90, vjust = .5, size=8, margin=margin(0,13,0,0)),
              legend.text=element_text(size=8, color='grey30'), 
              legend.title=element_text(size=8),
              legend.position=lp,
              legend.key.size=unit(.8,'line'),
              plot.title = element_text(size=8, hjust=0.5)) + 
        labs(color='tumor volume type') +
        scale_color_manual(values = wes_palette("Darjeeling1")[-4] )
      
      table <- cv[!(duplicated(cv$model)),]
      table <- table[, c('model','size','length','gap','support')]
      table <- melt(table, id='model')
      
      t <- ggplot(table, aes(x=model,y=variable, label=format(value))) +
        geom_text(size = 4.2, color='grey30', hjust=0.6) + theme_bw() + scale_y_discrete('cSPADE\nparameters') +  scale_x_discrete(expand=c(0.1,0)) +
        theme(panel.grid.major = element_blank(), legend.position = "none",
              panel.border = element_blank(),
              axis.text.x = element_blank(),
              axis.ticks = element_blank(),
              axis.text=element_text(8),
              axis.title.y = element_text(angle = 90, vjust = .5, size=8, margin=margin(0,11,0,5)) ) +
        theme(plot.margin = unit(c(0,.19, 0, 0), "lines")) + xlab(NULL) + ylab(NULL)

      a <- mmplot(g, t)
    }

    ```
    
1. **Parse the cross-validation log files**
    
    The following assumes a specific location and format of the log file (e.g., the rows skipped while reading `.log` files)
    ```{r}
    dataDir <- '~/gbm_spm_example_multi/logits'
    theseDirs <- list.dirs(path=dataDir, full.names=F, recursive=F)
    theseDirs
    
    # types and typenames are matched
    types <- c('percent','rateChange') # tumor volume types expected in log filename
    typenames <- c('percent change','rate change') # corresponding pretty names of tumor volume types
    months <- c('2-month','6-month','9-month', '12-month') # month names for classification labels
    
    cvs <- lapply(theseDirs, function(f) {
      cat(f, '\n')
      logFiles <- list.files(path=file.path(dataDir,f), pattern='\\.log$') # locate log files
      
      # extract results for each tumor volume type
      results <- lapply(seq(1,length(types)), function(i) {
        n <- logFiles[grepl(types[i],logFiles)] # read in a log file
        if (length(n) > 0) {
          log <- file.path(dataDir,f,n)
          con <- file(log)
          open(con)
          resultsCols <- read.table(con, skip=114, nrow=1, stringsAsFactors = F)  # expected results colnames
          r2 <- read.table(con, skip=0, nrow=1) # expected location of CV results for each classification
          r6 <- read.table(con, skip=25, nrow=1)
          r9 <- read.table(con, skip=25, nrow=1)
          r12 <- read.table(con, skip=25, nrow=1)
          close(con)
          
          r <- do.call('rbind', list(r2,r6,r9,r12)) # combine and format
          r <- r[,-1]
          colnames(r) <- c(resultsCols[1,])
          
          r$month <- months # attach additional info
          r$tType <- rep(typenames[i], nrow(r))
          r$model <- rep(f, nrow(r))
          temp <- do.call(rbind, lapply(r$model, function(n) regmatches(n, gregexpr('[0-9]+',n))[[1]][2:5] ) )
          colnames(temp) <- c('support', 'gap', 'length', 'size')
          temp <- as.data.frame(temp)
          temp$support <- paste0( '.', as.character(temp$support))
          r <- cbind(r, temp)
          r$name <- 'without'
          r
        } else {
          cat('\t ... file not found for ', typenames[i] , '\n')
        }
      })
      results <- do.call('rbind',results)
    })
    cvs <- do.call('rbind',cvs) 
    cat('duplicates?: ', which(duplicated((cvs))) )
    cvs[1:5,]
    ```

1. **Plot cross-validation results**

    ```{r, fig.width=3, fig.height=4, , warning = FALSE}
    # individual plots
    spmROC <- lapply(months, function(m) plotCVResults(getCVByMonth(cvResults=cvs, m, 'rocAUC'), 
                                             m, 
                                             '' ) )
    spmPR <- lapply(months, function(m) plotCVResults_PR(getCVByMonth(cvResults=cvs, m, 'prAUC'),
                                                m,
                                                '',
                                                c(.6,.8)) )
    ```
1. **Identify best model parameters from cross-validation**
    
    ```{r}
    # cv logit results
    bestLogitSpm <- lapply(months, function(m) {
      z <- cvs[which(cvs$month==m),]
      b <- z[which.max(z$rocAUC),]
      b <- b[,colnames(b) %in% c('rocAUC','rocAUCSD','tType','month','model', 'gap','length','name')]
      rownames(b) <- b$model
      b
    })
    bestLogitSpm <- do.call('rbind', bestLogitSpm)
    colnames(bestLogitSpm) <- c('rocAUC','rocAUCSD','month','name','model','maxgap','maxlength')
    bestLogitSpm$label <- getClassLabels() # assumes row order already matches
    bestLogitSpm$fitname <- c('fit2m','fit6m','fit9m','fit12m')
    bestLogitSpm
    ```
    
1. **Plot the cross-valiation results**
    
    There is only one approach explored in this example, i.e., approach 3, using temporal patterns and patient covariates to predict residual survival. The paper explores 2 other approaches.
    
    ```{r, fig.width=4, fig.height=3}
    comp <- bestLogitSpm[,c('rocAUC','rocAUCSD','month','name')]
    comp$lower <- comp$rocAUC - comp$rocAUCSD
    comp$higher <- comp$rocAUC + comp$rocAUCSD
    comp$approach <- 3
    comp <- do.call('rbind', list(comp))
    comp$approach <- as.factor(comp$approach)
    comp$approach <- factor(comp$approach, levels=levels(comp$approach), 
                            labels=c( 'temporal patterns\nwith covariates'))
    pd <- position_dodge(0.6) # move them .05 to the left and right
    
    comparedPlot <- ggplot(comp[!(comp$month=='12-month'),], aes(x=month, y=rocAUC, fill=factor(approach)) ) +
      geom_bar(stat='identity',position=position_dodge()) +
      scale_y_continuous(limits = c(.4,1), oob=rescale_none, breaks = seq(.4,1,0.1)) +
      theme_linedraw() +
      # ggtitle('') +
      perfTheme +
      geom_errorbar(aes(ymin=lower, ymax=higher), width=.2, position=position_dodge(.9)) +
      labs(fill='Best predictor\nin approach:') + xlab('Task') + ylab('AUC') +
      scale_fill_manual(values = wes_palette("Zissou1")[3:5] ) + 
      theme(legend.text.align=0, 
            legend.key.height=unit(1.5,'line'),
            legend.text = element_text(size=8),
            legend.title = element_text(size=8),
            axis.text=element_text(size=10),
            axis.title.y=element_text(size=10),
            axis.title.x=element_text(size=10)) +
      guides(fill = guide_legend(override.aes = list(size=6))) 
    print(comparedPlot)
    ```
    
1. **Retrain and apply to test set**

    A model is selected, retrained on the training partition, and applied to the testing partition for each prediction task. For the sake of this example, let's assume rate change as the tumor volume measurement was the best performer across all tasks.
    
    
    Prep for retraining:
    
    ```{r}
    seed <- 9
    metric <- 'rocAUC'
    expDir <- dirname(dataDir) 
    
    # data partitions
    dataPartitions <- file.path(expDir, 'train_test_partitions.rds')
    partitions <- readRDS(file=dataPartitions) 
    
    # unwanted features and labels that should not be in training data

    needToRemove <- c('id','iois','eventID', # remove ids
                      bestLogitSpm$label, # remove labels
                      'IDH1') # not interested

    # where to find the 'rate change' data for training
    selectedTType <- 'rate'
    selectedTTypeFN_spm <- 'featureVectors_rateChange.rds'
    selectedTTypeFN_cvLogit <- 'logits_CVresults_rateChange.rds'
    ```
    
    Get model performance results in retraining and testing:
    
    ```{r}
    retrainResults <- lapply(seq(1, nrow(bestLogitSpm)), function(i) {
      selected <- bestLogitSpm[i,]
      
      cat('\n....', selected$label ,' ...\n')
      
      # cv results
      cvLogitPath <- file.path(expDir,'logits', selected$model, selectedTTypeFN_cvLogit)
      cvResults <- readRDS(cvLogitPath)
      cvResults <- cvResults[[selected$fitname]]
      
      # get lambda from best cv results
      cvResults <- cvResults$avg.CVPerformance[which.max(cvResults$avg.CVPerformance$rocAUC),]
      cat('\n\n','Summary to repeated CV results (lambdas for retraining) ...\n\n')
      print(cvResults)
      
      
      # get data
      dataPath <- file.path(expDir,'spm', selected$model, selectedTTypeFN_spm)
      data <- readRDS(dataPath) # features and labels for each clinical visit
      names <- colnames(data)
      data <- data.frame(id=row.names(data),data)
      colnames(data) <- c('id',names)
      cat('... \n overall samples: ', nrow(data), '\n')
      
      data <- prepLaterality(data) 
      data <- prepLocation(data)
      
      cat('... \n removing clinical visits with no history of length maxgap:', selected$maxgap, '\n')
      data <- removeVisits(data = data, 
                           maxgap = as.numeric(as.character(selected$maxgap)), 
                           maxlength = as.numeric(as.character(selected$maxlength)), 
                           tType = selectedTType, 
                           save=F, 
                           outDir=outputDir)
      start.local <- Sys.time()
      cat('\n....', cvResults$lambda ,' ...\n')
      label <- selected$label
      
      fit <- retrainLogit(data = data[data$id %in% partitions$train,],  #training data
                          formula = as.formula(paste0(selected$label,'~.')), 
                          labelName = selected$label, 
                          lambda = cvResults$lambda,
                          lasso=TRUE, # using lamba
                          needToRemove=needToRemove, 
                          createModelMatrix=FALSE,
                          metric=metric, 
                          verbose=TRUE,
                          seed=seed)
      cat('\n'); print(Sys.time() - start.local)
      
      trainResults <-  getTrainResults(fit) 
      testResults <- getTestResults(fit = fit,
                                    data = data[data$id %in% partitions$test,], #testing data 
                                    labelName = selected$label, 
                                    formula = as.formula(paste0(selected$label,'~.')), 
                                    needToRemove=needToRemove, 
                                    createModelMatrix=FALSE)
      
      testResults$selectedThreshold <-  trainResults$selectedThreshold # use the threshold selected from train
      
      list(fit=fit, train=trainResults, test=testResults)
    })
    names(retrainResults) <- bestLogitSpm$fitname
    ```
    
    Contents of retrainResults:
    
    ```{r}
    names(retrainResults) # for each model
    names(retrainResults$fit2m) # for each prediction
    names(retrainResults$fit2m$train) # results in train
    names(retrainResults$fit2m$test) # results in test
    ```
    
    If we were to chose a threshold for classifiying 2-month residual survival, the results would be:
    
    ```{r}
    retrainResults$fit2m$train$selectedThreshold #train
    retrainResults$fit2m$test$selectedThreshold #test
    ```
    
1. **Plot final model's performance results in retraining and testing**
    
    Extract and structure the performance results:
    
    ```{r}
    rocsList <- lapply(retrainResults, function(fit) {list(train=fit$train$performance$roc,
                                                           test =fit$test$performance$roc )})
    aucsList <- lapply(retrainResults, function(fit) {list(train=fit$train$performance$aucROC,
                                                           test =fit$test$performance$aucROC )})
    prsList <- lapply(retrainResults, function(fit) {prs <- list(train=fit$train$performance$pr,
                                                                 test =fit$test$performance$pr )})
    praucsList <- lapply(retrainResults, function(fit) {list(train=fit$train$performance$aucPR,
                                                             test =fit$test$performance$aucPR)})
    names(rocsList) <- months # assumes months and order of retrainResults are the same
    names(aucsList) <- months
    names(prsList) <- months
    names(praucsList) <- months
    ```
    
    Plot ROC and PR of final model's test and train results
    
    ```{r, results='hide'}
    scaleColor <- scale_color_manual(values = c(wes_palette("Darjeeling1")[1], wes_palette('FantasticFox1')[3]) )
    lo <- c(2,3,4,1)

    finalModelROC <- plotMultiROC(fn = '',
                                  rocsList = rocsList[1:4],
                                  aucsList = aucsList[1:4],
                                  listOrder = lo,
                                  outDir = NULL,
                                  legendTitle = 'data\npartition', 
                                  printIt=F, 
                                  perfTheme = perfTheme,
                                  scaleX = scaleX,
                                  scaleY = scaleY,
                                  perfGuide = perfGuide) + 
      theme(legend.text.align=0) + 
      scaleColor 
    finalModelPR <- plotMultiPR(fn = '',
                                prsList = prsList[1:4], 
                                aucsList = praucsList[1:4], 
                                legendTitle = 'data\npartition',
                                outDir = NULL,
                                listOrder = lo,
                                printIt=F, 
                                perfTheme = perfTheme, 
                                scaleX = scaleX,
                                scaleY = scaleY, 
                                perfGuide = perfGuide) + 
      theme(legend.text.align=0) + 
      scaleColor
    ```
    ```{r, fig.width=9, fig.height=2.5, warning = FALSE}
    print(finalModelROC)
    print(finalModelPR)
    ```